Recursive partitioning of linear growth curve models

### 2022-10-25 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3230 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 5f4f1afb87aba5c7be522eac1dfddfb21339cf1f

\usepackage{Sweave}
\begin{document}
\input{_Partitioning_GCMs_with_GLMM_trees-concordance}
\maketitle

@@

@@

\begin{figure}[!ht]
\caption{Design of the subgroup-specific fixed effects in the simulation.}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-003}
\label{fig:design_tree}
\end{figure}

@@

\begin{figure}[!ht]
\caption{Tree size distributions for LM(M) trees.}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-009}
\end{subfigure}
{\footnotesize \textit{Note.} The $y$-axis is on the log scale and represents tree size. Grey lines indicate true number of splits (3 nodes). cluster = cluster-level covariances employed in parameter stability tests; ranef init = estimation initialized with the random effects. Distances on $y$-axis are on the log scale.}
\label{fig:LMM_sizes}
\end{figure}

Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach. The default fitting approach overfits, and implemented $> 3$ splits in most datasets. This overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits 2.87), which appears most effective for LMM trees (mean number of splits 3.02 for trees estimating only random intercepts; 2.96 for trees estimating random intercepts and slopes).

Initializing estimation with the random effects also effectively reduces tree size, but too strongly when random slopes are estimated (mean number of splits without random slopes 3.15; with random slopes 1.15). Heterogeneity related to partitioning variables measured at level 2 can either be captured and explained by the tree, or captured by the random effects. By initializing estimation with the random effects effects, this heterogeneity can no longer be captured by the tree; this likely worsens with more complex random-effects specification. Combining the use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding a similar and too high number of splits, irrespective of the random-effects specification (mean number of splits 4.1).

@@

% latex table generated in R 4.1.3 by xtable 1.8-4 package
% Wed Oct 26 00:12:30 2022
\begin{table}[ht]
\caption{Variables selected for the first split by each LM(M) tree approach.} 
\label{tab:first_splits}
\begin{tabular}{llrrrrr}
  \hline
Random effects & Fitting approach & $U_1$ & $U_2$ & $U_3$ & $U_4$ - $U_{25}$ & no split \\ 
  \hline
$\sigma_{b_0} = \sigma_{b_1} = 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} = \sigma_{b_1} = 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
   \hline
$\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & ran.eff. init. & 0.996 & 0.002 & 0.002 & 0.001 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & both & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
   \hline
$\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & ran.eff. init. & 0.499 & 0.007 & 0.006 & 0.003 & 0.485 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & both & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
   \hline \multicolumn{7}{l}{\footnotesize\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are}\\  \multicolumn{7}{l}{\footnotesize continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\sigma_{b_0}$ and $\sigma_{b_1}$} are\\  \multicolumn{7}{l}{\footnotesize the standard deviations of the random intercept and slope, respectively.}\\ \end{tabular}
\end{table}
\FloatBarrier

@@

\begin{figure}[!ht]
\caption{Tree size distributions for LM(M) trees with clustered covariances, SEM trees and LongCART.}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-014}
%\includegraphics{tree_size_plot.png}
\end{subfigure}
{\footnotesize \textit{Note.} Values on $x$-axis indicate whether random intercepts and/or slopes were estimated: no = random intercepts and slopes fixed to 0; int. = variance of random intercept freely estimated, variance of random slope fixed to 0; int. + slope = variance of random intercept and slope (and correlation between the two) freely estimated. Distances on $y$-axis are on the log scale. Grey lines indicate true number of splits (3 splits).}
\label{fig:tree_sizes}
\end{figure}

Figure~\ref{fig:tree_sizes} depicts tree size distributions for the different algorithms. LRT-based SEM trees appear somewhat overpowered with a mean number of splits of 3.61 ($SD = $1.71), while score-based SEM trees comprised the correct number of splits on average ($M = $3.09; $SD = $1.61). SEM trees appear more systematically affected by mis-specification of the random-effects structure than LMM trees and under-specification yields a higher number of splits. LongCART trees appear underpowered, with a mean number of splits of 2.01 ($SD = $1.45). This could be the result of under-specification of the random-effects structure: LongCART trees incorporate a random-intercept term only and this default cannot be overridden.

@@

@@

\begin{figure}[!b]
\caption{Tree sizes for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-028}
\end{subfigure}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-029}
\end{subfigure}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-030}
\end{subfigure}
\label{fig:application_sizes}
\end{figure}

>>>>

\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle

@@

<<eval=FALSE,echo=FALSE>>=
## If .Rnw will not compile bibliography, run tinytex on .tex file:
tinytex::latexmk("Partitioning_GCMs_with_GLMM_trees.tex")
@

@@

<<echo=FALSE>>=
library("partykit")
set.seed(42)
U1 <- sample(0:1, size = 1250, replace = TRUE) 
U2 <- round(rnorm(1250))
U3 <- round(rnorm(1250))
time <- rep(0:4, each = 250)
## left-most node: x1 == 0 & x2 <= 0
## intercept = -1, slope = -1
node3 <- as.numeric(U1 == 0 & U2 <= 0)
## left-middle node: x1 == 0 & x2 > 0
## intercept = -1, slope = 0
node4 <- as.numeric(U1 == 0 & U2 > 0)
## right-middle node: x1 == 1 & x3 =< 0
## intercept = 1, slope = 0
node6 <- as.numeric(U1 == 1 & U3 <= 0)
## right-most node: x1 == 1 & x3 > 0
## intercept = 1, slope = 1
node7 <- as.numeric(U1 == 1 & U3 > 0)
y <- (node3 * -1 - node3 * time) + (-node4) +
 (node6) + (node7 * 1 + node7 * time) 
data <- data.frame(U1 = as.factor(U1), U2, U3, time, y)   
tree <- lmtree(y ~ time | U1 + U2 + U3, data = data, maxdepth = 3)

fig4 <- party(
  partynode(1L,
            split = partysplit(1L, breaks = 1),
            kids = list(
              partynode(2L,
                        split = partysplit(2L, breaks = 0),
                        kids = list(partynode(3L, info = c(
                            expression(beta[j0] == '-1.0'),
                            expression(''),
                            expression(beta[j1] == '-1.0'))),
                          partynode(4L, info = c(
                            expression(beta[j0] == '-1.0'),
                            expression(''),
                            expression(beta[j1] == '0.0'))))),
              partynode(5L,
                        split = partysplit(3L, breaks = 0),
                        kids = list(
                          partynode(6L, info = c(
                            expression(beta[j0] == '1.0'),
                            expression(''),
                            expression(beta[j1] == '0.0'))),
                          partynode(7L, info = c(
                            expression(beta[j0] == '1.0'),
                            expression(''),
                            expression(beta[j1] == '1.0'))))))),
  data.frame(data)
)
#plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)

## Panel-combining function:
combine_panels <- function(panel1, panel2, party2 = NULL) {
  function(node) {
    nid <- id_node(node)
    pushViewport(viewport(
      layout = grid.layout(nrow = 2, ncol = 1, heights = c(1, 1)),
      name = paste("node_mob_mypanel", nid, sep = "")))
    grid.rect(gp = gpar(fill = "white", col = 0))
    pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
    panel1(node)
    popViewport()
    pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
    node2 <- if(is.null(party2)) node else node_party(party2[nid])
    panel2(node2)
    popViewport(2)
  }
}
@

\begin{figure}[!ht]
\caption{Design of the subgroup-specific fixed effects in the simulation.}
<<fig=T, height=5, width=5, echo=F>>=
#plot(tree, ip_args = list(pval = FALSE), gp = gpar(cex = .7),
#     tp_args = list(pointcol = "white", xlab = "time",
#                    mainlab = function(id, nobs) paste("node", id)))
plot(tree, tnex = 3.5, terminal_panel = combine_panels(
    panel1 = node_bivplot(tree, xlab = "time", pointcol = "white",
                        mainlab = function(id, nobs) paste("node", id)),
    panel2 = node_terminal(fig4, FUN = identity, height = 4, width = 12, id = FALSE),
    party2 = fig4), 
  gp = gpar(cex = .7), ip_args = list(pval = FALSE))
@
\label{fig:design_tree}
\end{figure}

@@

<<echo=FALSE>>=
load("tree_sizes.Rda")
tree_size <- data.frame(sizes, true = 7)
N <- c(100, 250)
sigma_int <- c(1, 2)
sigma_slope <- c(sqrt(.1), sqrt(.4))
p_noise <- c(5, 25)
rho <- c(0, .3)
design_matrix <- expand.grid(N = N, sigma_int = sigma_int,
                             sigma_slope = sigma_slope,
                             p_noise = p_noise, rho = rho)
design_matrix[] <- lapply(design_matrix, factor)
tree_sizes_long <- data.frame(stack(tree_size),
                              dataset_id = factor(rep(1:nrow(tree_size), 
                                                      times = ncol(tree_size))))
names(tree_sizes_long)[1:2] <- c("tree_size", "method_id") 
tree_sizes_long <- cbind(tree_sizes_long, design_matrix)
tree_sizes_long$method_id <- relevel(tree_sizes_long$method_id, ref = "true")
LMM_ids <- which(tree_sizes_long$method_id %in% c("true", "lm", "lm_c", "lmer", "lmer_c", 
                                                  "lmer_cr", "lmer_r", "lmer_s",
                                                  "lmer_s_c", "lmer_s_cr", "lmer_s_r"))
stdCoef <- function(object) {
  sdy <- sd(getME(object,"y"))
  sdx <- apply(getME(object,"X"), 2, sd)
  sc <- fixef(object)*sdx/sdy
  se.fixef <- coef(summary(object))[,"Std. Error"]
  se <- se.fixef*sdx/sdy
  return(data.frame(stdcoef=sc, stdse=se))
}
@

<<echo=FALSE, eval=FALSE>>=
library("lmerTest")
## LM(M) trees only
mod_LMM <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
  N + sigma_int + sigma_slope + p_noise + rho), 
  data = tree_sizes_long[LMM_ids, ])
#summary(mod_LMM)
#fixef(mod_LMM)[order(abs(fixef(mod_LMM)))]
coefs <- stdCoef(mod_LMM)[,1, drop = FALSE]^2

## "Variance explained" by N, sigma_n0, sigma_b1, p_noise and rho:
sum(coefs[grep("sigma_int", rownames(coefs)), ])
sum(coefs[grep("N", rownames(coefs)), ])
sum(coefs[grep("p_noise", rownames(coefs)), ])
sum(coefs[grep("sigma_slope", rownames(coefs)), ])
sum(coefs[grep("rho", rownames(coefs)), ])

## Note, above is not really variance explained, but some measure 
## of influence that does not add up to 1

## Variance explained by main effects:
coefs["N250", ] 
coefs["sigma_int2", ]
coefs["p_noise", ]
coefs["sigma_slope0.632455532033676", ]
coefs["rho0.3", ]

round(coefs[order(coefs$stdcoef, decreasing = TRUE), , drop = FALSE], digits = 3)

## Variance explained by random effects:
varco <- as.data.frame(VarCorr(mod_LMM))$vcov /
  var(tree_sizes_long[LMM_ids, "tree_size"])
varco
@

<<echo=FALSE>>=
library("lmerTest")
## Compare means between settings
tmp <-  tree_sizes_long[LMM_ids, ]
tmp$tree_size <- (tmp$tree_size-1)/2
tmp$default_ids <- tmp$method_id %in% c("lm", "lmer", "lmer_s")
tmp$cluster_ids <- tmp$method_id %in% c("lm_c", "lmer_c", "lmer_s_c")
tmp$ranefstart_ids <- tmp$method_id %in% c("lmer_r", "lmer_s_r")
tmp$both_ids <- tmp$method_id %in% c("lmer_cr", "lmer_s_cr")

M_default <- mean(tmp$tree_size[tmp$default_ids])
M_cluster <- mean(tmp$tree_size[tmp$cluster_ids])
M_ranefstart <- mean(tmp$tree_size[tmp$ranefstart_ids])
M_both <- mean(tmp$tree_size[tmp$both_ids])
@

<<eval=FALSE, echo=FALSE>>=
## t-tests for comparing no of splits with true number of splits
round(apply((sizes-1)/2, 2, function(x) t.test(x, mu = 3)$p.value), digits = 3)

## pairwise difference tests
pairwise.t.test(tmp$tree_size, tmp$method_id, 
               alternative = "two.sided", p.adjust = "bonferroni")
##          lmer    lmer_c  lmer_r  lmer_cr lmer_s  lmer_s_c lmer_s_r lmer_s_cr lm
## lmer_c    < 2e-16 -       -       -       -       -        -        -         -      
## lmer_r    < 2e-16 1.0000  -       -       -       -        -        -         -      
## lmer_cr   < 2e-16 < 2e-16 < 2e-16 -       -       -        -        -         -      
## lmer_s    0.0773  < 2e-16 < 2e-16 < 2e-16 -       -        -        -         -      
## lmer_s_c  < 2e-16 1.0000  0.0472  < 2e-16 < 2e-16 -        -        -         -      
## lmer_s_r  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16  -        -         -      
## lmer_s_cr < 2e-16 < 2e-16 < 2e-16 0.0027  < 2e-16 < 2e-16  < 2e-16  -         -      
## lm        1.0000  < 2e-16 < 2e-16 < 2e-16 1.0000  < 2e-16  < 2e-16  < 2e-16   -      
## lm_c      < 2e-16 2.4e-10 < 2e-16 < 2e-16 < 2e-16 4.6e-07  < 2e-16  < 2e-16   < 2e-16
@

<<echo=FALSE>>=
means <- tapply(tmp$tree_size, tmp$method_id, mean)
@

\begin{figure}[!ht]
\caption{Tree size distributions for LM(M) trees.}
\begin{subfigure}{1.2\textwidth}
<<echo=FALSE, fig=TRUE, height=3.5>>=
## Prepare data
tmp <-  tree_sizes_long
tmp$tree_size <- (tmp$tree_size-1)/2
tmp <- tmp[-which(tmp$method_id == "true") , ]
tmp$method_id <- factor(tmp$method_id)

## Create indicator for algorithm and random effects specification
tmp$ranef <- NA
tmp$ranef[tmp$method_id %in% c("lm", "lm_c")] <- "LM tree \n(no random effects)"
tmp$ranef[tmp$method_id %in% c("lmer", "lmer_c", "lmer_r", "lmer_cr")] <- "LMM tree \n(random intercept)"
tmp$ranef[tmp$method_id %in% c("lmer_s", "lmer_s_c", "lmer_s_r", "lmer_s_cr")] <- "LMM tree \n(random intercept + slope)"
tmp$ranef[tmp$method_id %in% c("LongCART")] <- "LongCART"
tmp$ranef[tmp$method_id %in% c("s_sem", "f_sem")] <- "SEM trees \n(no random effects)"
tmp$ranef[tmp$method_id %in% c("s_sem_i", "f_sem_i")] <- "SEM tree \n(random intercept)"
tmp$ranef[tmp$method_id %in% c("s_sem_s", "f_sem_s")] <- "SEM tree \n(random intercept + slope)"
tmp$ranef <- factor(tmp$ranef, 
                    levels = c("LM tree \n(no random effects)", "LMM tree \n(random intercept)", 
                               "LMM tree \n(random intercept + slope)", 
                               "SEM tree \n(no random effects)", "SEM tree \n(random intercept)", 
                               "SEM tree \n(random intercept + slope)"))

## Create indicator for additional settings
tmp$setting <- NA
tmp$setting[tmp$method_id %in% c("lmer", "lmer_s", "lm")] <- "default"
tmp$setting[tmp$method_id %in% c("lmer_c", "lmer_s_c", "lm_c")] <- "cluster"
tmp$setting[tmp$method_id %in% c("lmer_r", "lmer_s_r")] <- "ranef\ninit"
tmp$setting[tmp$method_id %in% c("lmer_cr", "lmer_s_cr")] <- "cluster +\nranef"
tmp$setting[tmp$method_id %in% c("LongCART")] <- "default"
tmp$setting[tmp$method_id %in% c("s_sem", "s_sem_i", "s_sem_s")] <- "score"
tmp$setting[tmp$method_id %in% c("f_sem", "f_sem_i", "f_sem_s")] <- "fair"
tmp$setting <- factor(tmp$setting, levels = c("cluster +\nranef", "cluster", 
                                              "default", "fair",
                                              "ranef\ninit", "score"))
levels(tmp$sigma_int) <- c("1", "4")
levels(tmp$sigma_slope) <- c("0.1", "0.4")
tmp$setting <- factor(tmp$setting, levels = c("default", "cluster",  "ranef\ninit", 
                       "cluster +\nranef", "fair", "score"))

## Create the plot
library("ggplot2")
library("gridExtra")
library("colorspace")
library("scales")

## LM(M)
LMM_ids <- which(tmp$method_id %in% c("lm", "lm_c", "lmer", "lmer_c", "lmer_cr", 
                                      "lmer_r", "lmer_s", "lmer_s_c", 
                                      "lmer_s_cr", "lmer_s_r"))
breaks <- c(0, 1, 3, 7, 20)
lim <- c(0, 25)
plot0 <- ggplot(tmp[LMM_ids, ]) +
  geom_boxplot(aes(x=setting, y=tree_size), 
               position=position_dodge(1), show.legend = FALSE, 
               alpha = 0.5, width = .6) + 
  scale_y_continuous(trans = pseudo_log_trans(sigma = 1, base = exp(1)), 
                     breaks = breaks, lim = lim) +
  facet_grid(~ranef, scales = "free", space = "free") +
  labs(x = "", y = "# of splits") + theme(legend.position = "none") +
  geom_hline(yintercept=3, col = "darkgrey")
plot0
@
\end{subfigure}
{\footnotesize \textit{Note.} The $y$-axis is on the log scale and represents tree size. Grey lines indicate true number of splits (3 nodes). cluster = cluster-level covariances employed in parameter stability tests; ranef init = estimation initialized with the random effects. Distances on $y$-axis are on the log scale.}
\label{fig:LMM_sizes}
\end{figure}

Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach. The default fitting approach overfits, and implemented $> 3$ splits in most datasets. This overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits \Sexpr{round(M_cluster, digits = 2)}), which appears most effective for LMM trees (mean number of splits \Sexpr{round(means["lmer_c"], digits = 2)} for trees estimating only random intercepts; \Sexpr{round(means["lmer_s_c"], digits = 2)} for trees estimating random intercepts and slopes).

Initializing estimation with the random effects also effectively reduces tree size, but too strongly when random slopes are estimated (mean number of splits without random slopes \Sexpr{round(means["lmer_r"], digits = 2)}; with random slopes \Sexpr{round(means["lmer_s_r"], digits = 2)}). Heterogeneity related to partitioning variables measured at level 2 can either be captured and explained by the tree, or captured by the random effects. By initializing estimation with the random effects effects, this heterogeneity can no longer be captured by the tree; this likely worsens with more complex random-effects specification. Combining the use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding a similar and too high number of splits, irrespective of the random-effects specification (mean number of splits \Sexpr{round(M_both, digits = 2)}).

@@

<<echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
library("xtable")
load(file = "splitvars.Rda")
splitvar <- apply(splits, 2, unlist)
rownames(splitvar) <- NULL

## LongCART registers best splitting variable if no split was implemented, overwrite
#table(splitvar[tree_size$Lcrt == 1, "Lcrt"])
# x1 x14 x16  x2 x28  x5  x9 
#607   1   1   1   2   1   1 
#
## Note: variable selection bias may not be a problem; power seems to be a problem
Lcrt_splitvars <- splitvar[, "Lcrt"] ## save for later use
splitvar[tree_size$Lcrt == 1, "Lcrt"] <- NA

## Get splitting proportions
func <- function(x) table(x, useNA = "ifany") / sum(table(x, useNA = "ifany"))
probs <- apply(splitvar, 2, func)
probs <- sapply(probs, function(x) {
  names(x)[is.na(names(x))] <- "no split"
  return(x)
})

## Have column for u1, u2, u3, u4-up, no split
splits <- matrix(sapply(probs, function(x) x["x1"]), ncol = 1,
                 dimnames = list(names(probs), "x1"))
for (u_set in list("x2","x3", paste0("x", 4:30), "no split")) {
  splits <- cbind(splits, sapply(probs, function(x) sum(x[names(x) %in% u_set])))
}
splits <- data.frame(splits[c(9:10, 1:8, 11:17), ]) ## reorder rows so LM tree is 1st
rows <- cbind(c("LM tree", " ", 
                "LMM tree", " ", " ", " ", " ", " ", " ", " ", 
                "SEM tree", " ", " ", " ", " ", " ", 
                "LongCART"),
                c("$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$"),
             c("", "clustered cov.", "", "clustered cov.", 
               "ran.eff. init.", "both", "", "clustered cov.", 
               "ran.eff. init.", "both",
               "LRT", "LRT", "LRT", "score-based", "score-based", "score-based", ""))
splits <- cbind(rows, splits)
colnames(splits) <- c("Algorithm", "Random effects", "Fitting approach", "$U_1$", 
                      "$U_2$", "$U_3$", "$U_4$ - $U_{25}$", "no split")

print(xtable(splits[1:10, -1], digits = 3,
             caption = "Variables selected for the first split by each LM(M) tree approach.", 
             label = "tab:first_splits"), 
      include.rownames=FALSE, hline.after = c(-1, 0, 2, 6), #table.placement = "tb",
      add.to.row = list(pos = list(10, 10, 10, 10), command = c(
        "\\hline",
        "\\multicolumn{7}{l}{\\footnotesize\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are}\\\\ ", 
        "\\multicolumn{7}{l}{\\footnotesize continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\\sigma_{b_0}$ and $\\sigma_{b_1}$} are\\\\ ",
        "\\multicolumn{7}{l}{\\footnotesize the standard deviations of the random intercept and slope, respectively.}\\\\ ")),
      latex.environment = "", caption.placement = "top",
      sanitize.text.function=function(x){x})
@

@@

<<echo=FALSE>>=
tmp2 <- tmp
LMM_c_ids <- which(tmp2$method_id %in% c("lm_c", "lmer_c", "lmer_s_c"))
SEM_ids <- which(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s",
                                      "s_sem", "s_sem_i", "s_sem_s"))
Lcrt_ids <- which(tmp2$method_id %in% "Lcrt")
tmp2 <-  tmp2[c(LMM_c_ids, SEM_ids, Lcrt_ids), ]
tmp2$method_id <- factor(tmp2$method_id)
@

<<echo=FALSE, eval=FALSE>>=
## SEM trees
SEM_ids <- which(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s",
                                      "s_sem", "s_sem_i", "s_sem_s"))
mod_SEM <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
  N + sigma_int + sigma_slope + p_noise + rho), 
  data = tmp2[SEM_ids, ])
summary(mod_SEM)
coefs <- stdCoef(mod_SEM)[,1, drop = FALSE]^2
r2_N <- sum(coefs[grep("N", rownames(coefs)), ])
r2_si <- sum(coefs[grep("sigma_int", rownames(coefs)), ])
r2_ss <- sum(coefs[grep("sigma_slope", rownames(coefs)), ])
r2_p <- sum(coefs[grep("p_noise", rownames(coefs)), ])
r2_rho <- sum(coefs[grep("rho", rownames(coefs)), ])

## LongCART
Lcrt_ids <- which(tmp2$method_id %in% "Lcrt")
mod <- lm(tree_size ~ N + sigma_int + sigma_slope + p_noise + rho, 
  data = tmp2[Lcrt_ids, ])
summary(mod)
coefs <- stdCoef(mod_LMM)[,1, drop = FALSE]^2
r2_N <- sum(coefs[grep("N", rownames(coefs)), ])
r2_si <- sum(coefs[grep("sigma_int", rownames(coefs)), ])
r2_ss <- sum(coefs[grep("sigma_slope", rownames(coefs)), ])
r2_p <- sum(coefs[grep("p_noise", rownames(coefs)), ])
r2_rho <- sum(coefs[grep("rho", rownames(coefs)), ])
@

<<eval=FALSE, echo=FALSE>>=
## LM(M) trees only
mod_all <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
  N + sigma_int + sigma_slope + p_noise + rho), data = tmp2)
#summary(mod_LMM)
#fixef(mod_LMM)[order(abs(fixef(mod_LMM)))]
coefs <- stdCoef(mod_all)[,1, drop = FALSE]^2

## "Variance explained" by N, sigma_n0, sigma_b1, p_noise and rho:
sum(coefs[grep("N", rownames(coefs)), ])
sum(coefs[grep("sigma_int", rownames(coefs)), ])
sum(coefs[grep("p_noise", rownames(coefs)), ])
sum(coefs[grep("rho", rownames(coefs)), ])
sum(coefs[grep("sigma_slope", rownames(coefs)), ])

## Note, above is not really variance explained, but some measure 
## of influence that does not add up to 1

## Variance explained by main effects:
coefs["N250", ] 
coefs["sigma_int2", ]
coefs["p_noise", ]
coefs["sigma_slope0.632455532033676", ]
coefs["rho0.3", ]

round(coefs[order(coefs$stdcoef, decreasing = TRUE), , drop = FALSE], digits = 3)

## Variance explained by random effects:
varco <- as.data.frame(VarCorr(mod_all))$vcov /
  var(tmp2[, "tree_size"])
varco
@

\begin{figure}[!ht]
\caption{Tree size distributions for LM(M) trees with clustered covariances, SEM trees and LongCART.}
\begin{subfigure}{1.2\textwidth}
<<echo=FALSE, fig=TRUE, height=3.5>>=
tmp2$method <- ifelse(
  tmp2$method_id %in% c("lm_c", "lmer_c", "lmer_s_c"), "LM(M) tree", 
      ifelse(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s"), "SEM tree (LRT-based)", 
             ifelse(tmp2$method_id %in% c("s_sem", "s_sem_i", "s_sem_s"),
                             "SEM tree (score-based)", "LongCART")))
tmp2$random <- ifelse(tmp2$method_id %in% c("lm_c", "f_sem", "s_sem"), "no", 
                      ifelse(tmp2$method_id %in% c("lmer_c", "f_sem_i", "s_sem_i", "Lcrt"),
                             "int.", "int. + \nslope"))
tmp2$random <- factor(tmp2$random, levels = c("no", "int.", "int. + \nslope"))
tmp2$method <- factor(tmp2$method, levels = c("LM(M) tree", "SEM tree (LRT-based)", 
                                              "SEM tree (score-based)", "LongCART"))
breaks <- c(0, 1, 3, 7, 15)
lim <- c(0, 15)
plot1 <- ggplot(tmp2) +
  geom_boxplot(aes(x=random, y=tree_size), alpha = .5, width = .6) + 
  geom_hline(yintercept=3, col = "darkgrey") +
  scale_y_continuous(trans = pseudo_log_trans(), 
                     breaks = breaks, lim = lim) +
  facet_grid(~ method, scales = "free", space = "free") +
  labs(x = "", y = "Tree size")
plot1
##ggsave(file = "tree_size_plot.png", height = 9.5, width=19, units = "cm")
@
%\includegraphics{tree_size_plot.png}
\end{subfigure}
{\footnotesize \textit{Note.} Values on $x$-axis indicate whether random intercepts and/or slopes were estimated: no = random intercepts and slopes fixed to 0; int. = variance of random intercept freely estimated, variance of random slope fixed to 0; int. + slope = variance of random intercept and slope (and correlation between the two) freely estimated. Distances on $y$-axis are on the log scale. Grey lines indicate true number of splits (3 splits).}
\label{fig:tree_sizes}
\end{figure}

Figure~\ref{fig:tree_sizes} depicts tree size distributions for the different algorithms. LRT-based SEM trees appear somewhat overpowered with a mean number of splits of \Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["SEM tree (LRT-based)"], digits = 2)} ($SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["SEM tree (LRT-based)"], digits = 2)}), while score-based SEM trees comprised the correct number of splits on average ($M = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["SEM tree (score-based)"], digits = 2)}; $SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["SEM tree (score-based)"], digits = 2)}). SEM trees appear more systematically affected by mis-specification of the random-effects structure than LMM trees and under-specification yields a higher number of splits. LongCART trees appear underpowered, with a mean number of splits of \Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["LongCART"], digits = 2)} ($SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["LongCART"], digits = 2)}). This could be the result of under-specification of the random-effects structure: LongCART trees incorporate a random-intercept term only and this default cannot be overridden.

<<eval=FALSE, echo=FALSE>>=
pairwise.t.test(tmp2$tree_size, tmp2$method_id, 
               alternative = "two.sided", p.adjust = "bonferroni")
# 	Pairwise comparisons using t tests with pooled SD 
# 
# data:  tmp2$tree_size and tmp2$method_id 
# 
#          lmer_c  lmer_s_c lm_c    f_sem   f_sem_i f_sem_s s_sem   s_sem_i s_sem_s
# lmer_s_c 1.00000 -        -       -       -       -       -       -       -      
# lm_c     < 2e-16 < 2e-16  -       -       -       -       -       -       -      
# f_sem    < 2e-16 < 2e-16  < 2e-16 -       -       -       -       -       -      
# f_sem_i  0.00074 2.5e-08  < 2e-16 < 2e-16 -       -       -       -       -      
# f_sem_s  1.00000 1.00000  < 2e-16 < 2e-16 9.6e-08 -       -       -       -      
# s_sem    < 2e-16 < 2e-16  < 2e-16 < 2e-16 < 2e-16 < 2e-16 -       -       -      
# s_sem_i  1.00000 1.00000  < 2e-16 < 2e-16 8.2e-09 1.00000 < 2e-16 -       -      
# s_sem_s  < 2e-16 < 2e-16  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 -      
# Lcrt     < 2e-16 < 2e-16  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 0.03511
@

@@

<<eval=FALSE, echo=FALSE>>=
## t-tests for comparing no of splits with true number of splits
round(apply((sizes-1)/2, 2, function(x) t.test(x, mu = 3)$p.value), digits = 3)
@

@@

<<eval=FALSE, echo=FALSE>>=
math_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = mathdata)
(as.data.frame(VarCorr(math_LMM))$vcov / var_math)
(as.data.frame(VarCorr(math_LMM))$vcov / as.data.frame(VarCorr(math_LMM))$vcov[4])

read_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = readdata)
(as.data.frame(VarCorr(read_LMM))$vcov / var_read)
(as.data.frame(VarCorr(read_LMM))$vcov / as.data.frame(VarCorr(read_LMM))$vcov[4])
